Finiteness in Cubical Type Theory

### 2020-06-30 sections/finiteness-predicates.tex
more work on bishop section

If that seems needlessly complex, don't worry: this definition is precisely
equivalent to the usual inductive one.
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/List/Base.tex]{list-def}
\end{agdalisting}
And this isn't some kind of hand-waving equivalence, either: since we working in
HoTT, we can (and do) prove that the two types are equal, allowing us to use one
or the other depending on whichever is more convenient, and
\(\AgdaFunction{subst}\) in the other representation without loss of generality.
That said, defining lists as containers will reveal several interesting
connections and proofs about split enumerability and the other predicates, 
so for the remainder of the paper whenever we say \AgdaDatatype{List} we will
mean Equation~\ref{list-def}.

Before we get to those interesting proofs, though, there are some other things
that need defining.
\AgdaDatatype{Fin}, for instance: \(\AgdaDatatype{Fin}\;n\) is the type of
natural numbers smaller than \(n\).
Its definition relies on a type for disjoint union,
\(\AgdaDatatype{\ensuremath{\uplus}}\).
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/Fin/Base.tex]{fin-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{disj-union}
  \end{agdalisting}%
  % It is also expressible with only \(\Sigma\):
  % \begin{agdalisting}
  %   \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-disj-union}
  % \end{agdalisting}
  % Although the inductive type definition is slightly more ergonomic.
\end{multicols} \noindent
And containers themselves, of course.
Containers are a well-studied topic in dependent type theory, with a rich
theory: we won't dive in to that here.
\begin{definition}[Containers] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S , P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{agdalisting} \label{container-interp}
    \ExecuteMetaData[agda/Container.tex]{container-interp}
  \end{agdalisting} \vspace{-\baselineskip}
\end{definition}

@@

To specify that a value must exist uniquely in HoTT we can use the concept of a
\emph{contraction}.

\begin{definition}[Contractions, Propositions and Sets] \label{homotopy-types}
  Types in HoTT and CuTT are not necessarily sets, as they are in MLTT.
  Some have higher homotopies (paths which aren't unique).
  We actually have a hierarchy of complexity of structure of path spaces in
  types, starting with the contractions \cite[definition 3.11.1]{hottbook}, then
  the mere propositions \cite[definition 3.3.1]{hottbook}, and the sets
  \cite[definition 3.1.1]{hottbook}.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{hlevels}
  \end{agdalisting}
  First we have the contractions: those types equivalent to
  \AgdaDatatype{\ensuremath{\top}}.
  Then the propositions: those types for which every value is equal to every
  other.
  Finally the sets: those types for which every \emph{equality} is equal to
  every other.

  Often these types are defined inductively: going from contractions upwards,
  every new level has paths in the level before (i.e. the paths of a proposition
  are all contractions).
  We can also proceed further upwards, into non-set types.
\end{definition}
\begin{definition}[Unique Membership] \label{uniq-memb-def}
  Unique list membership is defined in terms of list membership: it is a
  contraction of it.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Container/Membership.tex]{uniq-memb-def}
  \end{agdalisting}
\end{definition}

With this we can define manifest Bishop finiteness:
\begin{definition}[Manifest Bishop Finiteness]  \label{bish-def}
  A type is manifest Bishop finite if there exists a list which contains each
  value in the type once.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop/Container.tex]{bish-def}
  \end{agdalisting}
\end{definition}
The only difference between manifest Bishop finiteness and split enumerability
is the membership term: here we require unique membership (\(\inunique\)),
rather than simple membership (\(\in\)).

We use the word ``manifest'' here to distinguish from another common
interpretation of Bishop finiteness, which we have called cardinal finiteness in
this paper.
The ``manifest'' refers to the fact that we have a concrete, non-truncated list
of the elements in the proof.

\subsection{The Relationship Between Manifest Bishop Finiteness and Split
  Enumerability}
While manifest Bishop finiteness might seem stronger than split enumerability,
it turns out this is not the case.
Both predicates imply the other.
\begin{lemma} \label{manifest-bishop-to-split-enum}
  Any manifest Bishop finite type is split enumerable.
\end{lemma}
\begin{proof}
  To construct a proof of split enumerability from one of manifest Bishop
  finiteness, it suffices to convert a proof of \(x \inunique \mathit{xs}\) to
  one of \(x \in \mathit{xs}\), for all \(x\) and \(\mathit{xs}\).
  Since \(\inunique\) is defined as a contraction of \(\in\), such a conversion
  is simply the \(\text{fst}\) function.
\end{proof}

\begin{lemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{lemma}
This proof takes significantly more work.
The ``unique membership'' condition in
\(\AgdaDatatype{\ensuremath{\mathcal{B}}}\) means that we are not permitted
duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\) proof: we can do this using the decidable
equality provided by \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\) (lemma~\ref{split-enum-discrete}).
From there, we need to show that the membership proof carries over
appropriately.
\todo{Provide more info on this proof?}

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power.
\subsection{From Manifest Bishop Finiteness to Equivalence}
We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.

\begin{definition}[Equivalences] \label{equivalences}
  We will take contractible maps \cite[definition 4.4.1]{hottbook} as our
  ``default'' definition of equivalences.
  \begin{agdalisting} \label{is-equiv-def}
    \ExecuteMetaData[agda/Snippets/Equivalence.tex]{is-equiv-def}
  \end{agdalisting} \vspace{-.5\baselineskip}
  \begin{agdalisting} \label{equiv-def}
    \ExecuteMetaData[agda/Snippets/Equivalence.tex]{equiv-def}
  \end{agdalisting}
\end{definition}
\begin{lemma} \label{bishop-equiv}
  Manifest bishop finiteness is equivalent to an equivalence to a finite prefix
  of the natural numbers.
  \begin{equation}
  \end{equation}
\end{lemma}
\begin{proof} \todo{Convert to Agda}
  \begin{align*}
     \AgdaDatatype{\ensuremath{\mathcal{B}}}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \inunique \mathit{xs}
    && \text{def.~\ref{bish-def} }(\AgdaDatatype{\ensuremath{\mathcal{B}}})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(x \in \mathit{xs})
    && \text{eqn.~\ref{uniq-memb-def} } (\inunique)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(\text{fib}_{\text{snd}(\mathit{xs})}(x))
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{is-equiv-def} (isEquiv)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \AgdaDatatype{Fin} \rrbracket (A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{list-def} } (\AgdaDatatype{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \AgdaDatatype{Fin}\;n) , A) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \AgdaDatatype{Fin}\;n \rightarrow A) , \text{isEquiv}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \AgdaDatatype{Fin}\;n \simeq A )
    && \text{eqn.~\ref{equiv-def} } (\simeq) \; \qedhere
  \end{align*}
\end{proof}
This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}

>>>>

% \begin{wrapfigure}[5]{r}{.4\linewidth} \vspace{-1.6\baselineskip}
%   \ExecuteMetaData[agda/Data/List/Base.tex]{list-def} \vspace{-\baselineskip}
%   \caption{The Usual Inductive Definition}
% \end{wrapfigure}

If that seems needlessly complex, don't worry: this definition is precisely
equivalent to the usual inductive one.
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/List/Base.tex]{list-def}
\end{agdalisting}
And this isn't some kind of hand-waving equivalence, either: since we are
working in HoTT, we can (and do) prove that the two types are equal, allowing us
to use one or the other depending on whichever is more convenient, and
\(\AgdaFunction{subst}\) in the other representation without loss of generality.
That said, defining lists as containers will reveal several interesting
connections and proofs about split enumerability and the other predicates, 
so for the remainder of the paper whenever we say \AgdaDatatype{List} we will
mean Equation~\ref{list-def}.

Before we get to those interesting proofs, though, there are some other things
that need defining.
\AgdaDatatype{Fin}, for instance: \(\AgdaDatatype{Fin}\;n\) is the type of
natural numbers smaller than \(n\).
Its definition relies on a type for disjoint union,
\(\AgdaDatatype{\ensuremath{\uplus}}\).
\begin{multicols}{2} \null \vfill
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/Fin/Base.tex]{fin-def}
  \end{agdalisting} \vfill \null \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Introduction.tex]{disj-union}
  \end{agdalisting}%
  % It is also expressible with only \(\Sigma\):
  % \begin{agdalisting}
  %   \ExecuteMetaData[agda/Snippets/Introduction.tex]{sigma-disj-union}
  % \end{agdalisting}
  % Although the inductive type definition is slightly more ergonomic.
\end{multicols} \noindent
And containers themselves, of course.
Containers are a well-studied topic in dependent type theory, with a rich
theory: we won't dive in to that here.
\begin{definition}[Containers] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S , P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{agdalisting} \label{container-interp}
    \ExecuteMetaData[agda/Container.tex]{container-interp}
  \end{agdalisting} \vspace{-\baselineskip}
\end{definition}

@@

To specify that a value must exist uniquely in HoTT we can use the concept of a
\emph{contraction} \cite[definition 3.11.1]{hottbook}.
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Introduction.tex]{isContr}
\end{agdalisting}
A contraction is a type with the least possible amount of information: it
represents the tautologies.
All contractions are isomorphic to \AgdaDatatype{\ensuremath{\top}}. \todo{We'll
now need to define propositions and sets later on}

By saying that a proof of membership is a contraction, we are saying that it
must be \emph{unique}.
\begin{agdalisting} \label{uniq-memb-def}
  \ExecuteMetaData[agda/Container/Membership.tex]{uniq-memb-def}
\end{agdalisting}
Now a proof of \(x\;\AgdaFunction{\ensuremath{\in}!}\;\mathit{xs}\) means that
\(x\) is not just in \(\mathit{xs}\), but it appears there \emph{only once}.

% \begin{definition}[Contractions, Propositions and Sets] \label{homotopy-types}
%   Types in HoTT and CuTT are not necessarily sets, as they are in MLTT.
%   Some have higher homotopies (paths which aren't unique).
%   We actually have a hierarchy of complexity of structure of path spaces in
%   types, starting with the contractions , then
%   the mere propositions \cite[definition 3.3.1]{hottbook}, and the sets
%   \cite[definition 3.1.1]{hottbook}.
%   First we have the contractions: those types equivalent to
%   \AgdaDatatype{\ensuremath{\top}}.
%   Then the propositions: those types for which every value is equal to every
%   other.
%   Finally the sets: those types for which every \emph{equality} is equal to
%   every other.

%   Often these types are defined inductively: going from contractions upwards,
%   every new level has paths in the level before (i.e. the paths of a proposition
%   are all contractions).
%   We can also proceed further upwards, into non-set types.
% \end{definition}
% \begin{definition}[Unique Membership] \label{uniq-memb-def}
%   Unique list membership is defined in terms of list membership: it is a
%   contraction of it.
% \end{definition}

With this we can define manifest Bishop finiteness:
\begin{definition}[Manifest Bishop Finiteness]  \label{bish-def}
  A type is manifest Bishop finite if there exists a list which contains each
  value in the type once.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop/Container.tex]{bish-def}
  \end{agdalisting}
  The only difference between manifest Bishop finiteness and split enumerability
  is the membership term: here we require unique membership
  (\AgdaFunction{\ensuremath{\in!}}), rather than simple membership
  (\AgdaFunction{\(\in\)}).
\end{definition}

We use the word ``manifest'' here to distinguish from another common
interpretation of Bishop finiteness, which we have called cardinal finiteness in
this paper: this version of the proof is ``manifest'' because we have a
concrete, non-truncated list of the elements in the proof.

\subsection{The Relationship Between Manifest Bishop Finiteness and Split
  Enumerability}
While manifest Bishop finiteness might seem stronger than split enumerability,
it turns out this is not the case.
Both predicates imply the other.

Going from manifest Bishop finiteness is relatively straightforward:
to construct a proof of split enumerability from one of manifest Bishop
finiteness, it suffices to convert a proof of \(x \inunique \mathit{xs}\) to
one of \(x \in \mathit{xs}\), for all \(x\) and \(\mathit{xs}\).
Since \(\inunique\) is defined as a contraction of \(\in\), such a conversion
is simply the \(\text{fst}\) function.

Going the other direction takes significantly more work.
\begin{lemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{lemma}
We will only sketch the proof here:
the ``unique membership'' condition in
\(\AgdaDatatype{\ensuremath{\mathcal{B}}}\) means that we are not permitted
duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\) proof: we can do this using the decidable
equality provided by \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\) (lemma~\ref{split-enum-discrete}).
From there, we need to show that the membership proof carries over
appropriately.
\todo{Provide more info on this proof?}

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power.
\subsection{From Manifest Bishop Finiteness to Equivalence}

We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.
\begin{multicols}{2}
  \begin{agdalisting} \label{is-equiv-def}
    \ExecuteMetaData[agda/Snippets/Equivalence.tex]{is-equiv-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting} \label{equiv-def}
    \ExecuteMetaData[agda/Snippets/Equivalence.tex]{equiv-def}
  \end{agdalisting}
\end{multicols}

\begin{definition}[Equivalences] \label{equivalences}
  We will take contractible maps \cite[definition 4.4.1]{hottbook} as our
  ``default'' definition of equivalences.
\end{definition}
\begin{lemma} \label{bishop-equiv}
  Manifest bishop finiteness is equivalent to an equivalence to a finite prefix
  of the natural numbers.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-is-equiv-type}
  \end{agdalisting}
\end{lemma}

\begin{proof} \let\qed\relax \vspace{-1\baselineskip}
  \begin{minipage}[t]{.7\textwidth}\vspace{-\baselineskip}
    \begin{agdalisting*}
      \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-is-equiv}
    \end{agdalisting*}
  \end{minipage}
  \begin{minipage}[t]{.19\textwidth} \setstretch{1.1}

    Def.~\ref{bish-def} (\AgdaDatatype{\ensuremath{\mathcal{B}}}) \\
    Eqn.~\ref{uniq-memb-def} (\AgdaDatatype{\ensuremath{\in!}}) \\
    Eqn.~\ref{container-membership} (\AgdaDatatype{\ensuremath{\in}}) \\
    Eqn.~\ref{is-equiv-def}  \\
    Eqn.~\ref{list-def} (\(\AgdaDatatype{List}\)) \\
    Eqn.~\ref{container-interp}  \\
    Reassociation \\
    Eqn.~\ref{equiv-def}
  \end{minipage}
\end{proof}

% \begin{proof} \todo{Convert to Agda}
%   \begin{align*}
%      \AgdaDatatype{\ensuremath{\mathcal{B}}}(A) &
%     \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \inunique \mathit{xs}
%     && \text{def.~\ref{bish-def} }(\AgdaDatatype{\ensuremath{\mathcal{B}}})
%     \\
%     & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(x \in \mathit{xs})
%     && \text{eqn.~\ref{uniq-memb-def} } (\inunique)
%     \\
%     & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(\text{fib}_{\text{snd}(\mathit{xs})}(x))
%     && \text{eqn.~\ref{container-membership} } (\in)
%     \\
%     & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
%     && \text{eqn.~\ref{is-equiv-def} (isEquiv)}
%     \\
%     & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \AgdaDatatype{Fin} \rrbracket (A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
%     && \text{def.~\ref{list-def} } (\AgdaDatatype{List})
%     \\
%     & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \AgdaDatatype{Fin}\;n) , A) , \text{isEquiv}(\text{snd}(\mathit{xs}))
%     && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
%     \\
%     & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \AgdaDatatype{Fin}\;n \rightarrow A) , \text{isEquiv}(f)
%     && \text{Reassociation of } \Sigma
%     \\
%     & \simeq \Sigma (n : \mathbb{N}) , ( \AgdaDatatype{Fin}\;n \simeq A )
%     && \text{eqn.~\ref{equiv-def} } (\simeq) \; \qedhere
%   \end{align*}
% \end{proof}
This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}
